### 2020-12-16 latex/sections/results.tex
latexing
yngvi+james comments fixed

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsrandom.png}
        \end{center}
        \caption{Winrate vs random agent}
        \label{fig:winratevsrandom}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsmcts.png}
        \end{center}
        \caption{Winrate vs MCTS agent}
        \label{fig:winratevsmcts}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsneuralnetwork.png}
        \end{center}
        \caption{Winrate vs neural network trained for 150 generations}
        \label{fig:winratevsneuralnetwork}
    \end{small}
\end{figure}

Examining the neural network is only interesting if the neural network can effectively play the game. We first examine the history neural network playing against an agent that only takes random moves, on every $10$ generations. The win rate over generations can be seen in Figure \ref{fig:winratevsrandom}. It is clear that very early on in the training process, almost immediately after generation $10$ our agent performs nearly perfectly against the random agent. This implies that the agent does indeed learn some strategy in the game. The next agent we tried the neural network against, was an agent that does regular Monte-Carlo tree search on the game space with $100$ iterations of rollout for each move. A graph showing the win rate over generations is depicted in Figure \ref{fig:winratevsmcts}. Again, our agent quickly learns some strategy and can win often, although doesn't achieve a perfect win rate after $300$ generations. The last agent we tested against was the neural network itself, although only trained to $150$ generations. The graph in Figure \ref{fig:winratevsneuralnetwork} shows win rate. As expected the agent performs poorly until it has trained for more than $150$ generations.

These results imply that our agent certainly understands how to play the game of breakthrough to a certain level. However, an intermediate level human player would most often be able to win the agent, based on the author's experience playing against it. Note, that the main focus of this was not to create an expert level agent. Rather to, gauge into the concepts the agent learns as described in the next section.

\section{Evaluating the improvements of the Neural Network over generations}

@@

We test four concepts \textit{Material Advantage}, \textit{Aggressiveness}, \textit{Unity}, and \textit{Lorentz-Horey}.

\subsection{Description of concepts}

@@

The first concept we examine is Material Advantage which conveys the idea of having more pawns than your opponent. When faced with a board it is difficult to argue for whether having a single pawn up on your opponent constitues as Material Advantage or three. In order to an appropriate value for this case we used the neural network that had been trained for $300$ generations to generate $20000$ unique states. The distribution of these states' difference of pawns can be seen in Figure \ref{fig:distnumberadv}. From examining the distribution and examining samples we selected the breakpoint of $\geq 2$, meaning that if you have $2$ pieces on your opponent you are in a state that has this concept. From the $20000$ states only $5.5\%$ of states have this concept.

@@

The next concept is Aggressiveness, which describes how much closer your furthest piece is to the opponents edge than your opponents furthest piece to your edge. This is then the minimum amount of how many moves it will take you to win the game. A distribution of these values can be seen in Figure \ref{fig:distaggressiveness}. We examined the distribution, and sampled states from various breaking points and decided on the value of $\geq 1$. That is if the value of a state's furthest piece difference is greater than or equal to $1$ that state has the concept of Aggressiveness. From the dataset $18.3\%$ of states have this concept.

@@

\begin{table}[]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        5  & 15 & 5  & 5  & 15 & 5  \\\hline
        2  & 3  & 3  & 3  & 3  & 2  \\\hline
        4  & 6  & 6  & 6  & 6  & 4  \\\hline
        7  & 10 & 10 & 10 & 10 & 7  \\\hline
        11 & 15 & 15 & 15 & 15 & 11 \\\hline
        21 & 21 & 21 & 21 & 21 & 21 \\\hline
    \end{tabular}
    \caption{Lorentz-Horey cell values, from white players point of view}
    \label{table:lorentzcell}
\end{table}

The last concept we examined was the Lorentz-Horey. This concept is a popular heuristic in Breakthrough defined in the paper by Lorentz \& Horey \cite{lorentz:heuristic}. For this heuristic, we give each cell on the board a point value. For a given player the heuristic is then the sum of the cell values where they have pawns. The cell values are shown in Table \ref{table:lorentzcell}. The matrix shown is flipped from the black player's point of view. The values are selected in such a way that as your pawn approach the opponent's side their value increases, having pieces on the edge isn't optimal, as such a pawn will only be able to capture in one direction and can not escape capture as easily. To select the breakpoint for Lorentz-Horey heuristic, we generate a distribution of the difference in Lorentz-Horey Score over our dataset. The resulting distribution is shown in Figure \ref{fig:distlorentz}. Our examination lead us to select the value of $5$, as a breakpoint indicating that a state has the concept of Lorentz-Horey. Out of our dataset of 20,000 states $29.4\%$ of states have this concept.

@@

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/number_pawns_trend}
    \caption{Percentage of selected states containing the HLC Material Advantage}
    \label{fig:numberadvantage}
\end{figure}

@@

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/most_advanced_trend.png}
    \caption{Percentage of selected states containing the HLC aggressiveness}
    \label{fig:aggressiveness}
\end{figure}

From Figure \ref{fig:aggressiveness}. we can see that the aggressiveness HLC is a growing factor over as the neural network is trained. As a piece can always move forward, if your opponent does not capture that piece, this concept relates to how many moves it will take for you to win. Generally, when your opponent is not playing well this concept does well, and quickly becomes a bad heuristic in classical search algorithms.

@@

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/unity_trend.png}
    \caption{Percentage of selected states containing the HLC unity}
    \label{fig:unity}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/lorentz_horey_trend.png}
        \end{center}
        \caption{Percentage of selected states containing the HLC Lorentz-Horey}
        \label{fig:lorentzheuristic}
    \end{small}
\end{figure}

The literature of Breakthrough implies that being patient and waiting until your opponent makes a mistake is generally the preferred strategy for winning.\cite{lorentz:heuristic} Examining Figure \ref{fig:lorentzheuristic}, we can see that the trend of using this HLC increases over time. As this heuristic is the one most often cited as a successful strategy in Breakthrough, it is interesting to see that our Breakthrough agent is gradually learning a similar concept.

>>>>

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsrandom.png}
        \end{center}
        \caption{Winrate vs. random agent}
        \label{fig:winratevsrandom}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsmcts.png}
        \end{center}
        \caption{Winrate vs. MCTS agent}
        \label{fig:winratevsmcts}
    \end{small}
\end{figure}

\begin{figure}
    \begin{small}
        \begin{center}
            \includegraphics[width=0.7\textwidth]{graphics/winratevsneuralnetwork.png}
        \end{center}
        \caption{Winrate vs. neural network trained for 150 generations}
        \label{fig:winratevsneuralnetwork}
    \end{small}
\end{figure}

Examining the neural network is only interesting if the neural network can effectively play the game. We first examine the history of the neural network playing against an agent that only takes random moves. The win rate over every $10$ generations can be seen in Figure \ref{fig:winratevsrandom}. It is clear that very early on in the training process, almost immediately after generation $10$, our agent performs nearly perfectly against the random agent. This implies that the agent does indeed learn some strategy in the game. The next agent we tried the neural network against was an agent that does regular Monte-Carlo tree search on the game space with $100$ iterations of random UCT rollout for each move. The MCTS agent has a $c_{uct}$ value of $0.9$. A graph showing the win rate over generations is depicted in Figure \ref{fig:winratevsmcts}. Again, our agent quickly learns some strategy and can win often, although not achieving a perfect win rate even after $300$ generations. The last agent we tested against was the neural network itself, although only trained to $150$ generations. The graph in Figure \ref{fig:winratevsneuralnetwork} shows win rate. As one would expect, the agent that has trained for less time performs poorly until it has trained for more than $150$ generations.

These results imply that our agent certainly understands how to play the game of Breakthrough to a certain level. However, an intermediate level human player would most often be able to beat the agent, based on the author's experience playing against it. Note that the main focus of this was not to create an expert level agent. Rather, to gauge into the concepts the agent learns as described in the next section.

\section{Evaluating the Improvements of the Neural Network over Generations}

@@

We test four concepts \textit{Material Advantage}, \textit{Aggressiveness}, \textit{Unity}, and \textit{Lorentz-Horey score}.

\subsection{Description of Concepts}

@@

The first concept we examine is Material Advantage which conveys the idea of having more pawns than the opponent. When faced with a board it is difficult to argue for whether having a single pawn or three up on your opponent constitues as significant Material Advantage. In order to select an appropriate value for this case we used the neural network that had been trained for $300$ generations to generate $20{,}000$ unique states. The distribution of these states' difference of pawns can be seen in Figure \ref{fig:distnumberadv}. From examining the distribution and examining samples we selected the breakpoint of $\geq 2$, meaning that if you have $2$ pieces on your opponent you are in a state that has this concept. From the $20{,}000$ states only $5.5\%$ of states have this concept.

@@

The next concept is Aggressiveness, which describes how much closer your furthest piece is to the opponent's edge than your opponent's furthest piece to your edge. This is the minimum amount of how many moves it will take you to possibly win the game. A distribution of these values can be seen in Figure \ref{fig:distaggressiveness}. We examined the distribution, and sampled states from various breaking points and decided on the value of $\geq 1$. That is, if the value of a state's furthest piece difference is greater than or equal to $1$ that state has the concept of Aggressiveness. From the dataset $18.3\%$ of states have this concept.

@@

\begin{table}[]
    \centering
    \caption{Lorentz-Horey cell values, from white players point of view}
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        21 & 21 & 21 & 21 & 21 & 21 \\\hline
        11 & 15 & 15 & 15 & 15 & 11 \\\hline
        7  & 10 & 10 & 10 & 10 & 7  \\\hline
        4  & 6  & 6  & 6  & 6  & 4  \\\hline
        2  & 3  & 3  & 3  & 3  & 2  \\\hline
        5  & 15 & 5  & 5  & 15 & 5  \\\hline
    \end{tabular}
    \label{table:lorentzcell}
\end{table}

\label{section:lorentzhoreyref}

The last concept we examined was the Lorentz-Horey score. This concept is a popular heuristic in Breakthrough defined in the paper by Lorentz \& Horey \cite{lorentz:heuristic}. For this heuristic, we give each cell on the board a point value. For a given player the heuristic is then the sum of the cell values where they have pawns. The cell values are shown in Table \ref{table:lorentzcell}. The matrix shown is flipped from the black player's point of view. The values are selected in such a way that as your pawns approach the opponent's side their value increases, having pieces on the edge isn't optimal, as such a pawn will only be able to capture in one direction and can not escape capture as easily. To select the breakpoint for Lorentz-Horey heuristic, we generate a distribution of the difference in Lorentz-Horey Score over our dataset. The resulting distribution is shown in Figure \ref{fig:distlorentz}. Our examination lead us to select the value of $5$, as a breakpoint indicating that a state has the concept of Lorentz-Horey. Out of our dataset of 20,000 states $29.4\%$ of states have this concept.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/number_pawns_trend}
    \caption{Percentage of selected states containing the HLC Material Advantage}
    \label{fig:numberadvantage}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/most_advanced_trend.png}
    \caption{Percentage of selected states containing the HLC aggressiveness}
    \label{fig:aggressiveness}
\end{figure}

From Figure \ref{fig:aggressiveness}, we can see that the aggressiveness HLC is a growing factor over as the neural network is trained. As a piece can always move forward, if your opponent does not capture that piece, this concept relates to how many moves it will take for you to win. Generally, when your opponent is not playing well this concept does well, and quickly becomes a bad heuristic in classical search algorithms.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/unity_trend.png}
    \caption{Percentage of selected states containing the HLC unity}
    \label{fig:unity}
\end{figure}

Examining Figure \ref{fig:unity}, we see that the Unity HLC is steady, but then it drops as the network is trained. As this is considered a decent strategy in Breakthrough this is unexpected. However, this concept become increasingly unreliable as you lose pieces because it is the average of your remaining pieces distance to their center.

\subsection{Lorentz-Horey Score}

\begin{figure}[h]
    \begin{small}
        \begin{center}
            \includegraphics[width=0.95\textwidth]{graphics/lorentz_horey_trend.png}
        \end{center}
        \caption{Percentage of selected states containing the HLC Lorentz-Horey}
        \label{fig:lorentzheuristic}
    \end{small}
\end{figure}

Examining Figure \ref{fig:lorentzheuristic}, we can see a trend where as the neural network is trained its use of this HLC increases over time. As this heuristic is the one most often cited as a successful strategy in Breakthrough, it is interesting to see that our Breakthrough agent is gradually learning a similar concept.
